Architectures and Open pNets

### 2018-02-13 2018-Arch-pNets/arch-pNets.tex
Moved notations to a new Section 2 'Preliminaries'; revised and added data and valuations.

\newcommand{\noteSB}[2][color=green!40, size=\tiny]{\todo[#1]{{#2}\Simon}}
\newcommand{\noteSBin}[2][inline,color=green!40]{\todo[#1]{{#2}\Simon}}
\newcommand{\todoSB}[2][color=green!40, size=\tiny]{\todo[#1]{\textbf{To-do Simon:} {#2}}}
\newcommand{\todoSBin}[2][inline,color=green!40]{\todo[#1]{\textbf{To-do Simon: } {#2}}}
\newcommand{\Ludo}{\\\hfill\mdash Ludo}
\newcommand{\noteLH}[2][color=blue!40, size=\tiny]{\todo[#1]{{#2}\Ludo}}
\newcommand{\noteLHin}[2][inline,color=blue!40]{\todo[#1]{{#2}\Ludo}}
\newcommand{\todoLH}[2][color=blue!40, size=\tiny]{\todo[#1]{\textbf{To-do Ludo:} {#2}}}
\newcommand{\todoLHin}[2][inline,color=blue!40]{\todo[#1]{\textbf{To-do Ludo: } {#2}}}

@@

@@

\newcommand{\cA}{\ensuremath{\mathcal{A}}}
\newcommand{\bB}{\ensuremath{\mathbf{B}}}
\newcommand{\cB}{\ensuremath{\mathcal{B}}}
\newcommand{\sB}{\ensuremath{\mathbb{B}}}
\newcommand{\cC}{\ensuremath{\mathcal{C}}}
\newcommand{\sC}{\ensuremath{\mathbb{C}}}
\newcommand{\cD}{\ensuremath{\mathcal{D}}}
\newcommand{\fD}{\ensuremath{\mathsf{D}}}
\newcommand{\sD}{\ensuremath{\mathbb{D}}}
\newcommand{\cE}{\ensuremath{\mathcal{E}}}
\newcommand{\sE}{\ensuremath{\mathbb{E}}}
\newcommand{\cF}{\ensuremath{\mathcal{F}}}
\newcommand{\cG}{\ensuremath{\mathcal{G}}}
\newcommand{\cH}{\ensuremath{\mathcal{H}}}
\newcommand{\cI}{\ensuremath{\mathcal{I}}}
\newcommand{\sI}{\ensuremath{\mathbb{I}}}
\newcommand{\cM}{\ensuremath{\mathcal{M}}}
\newcommand{\cN}{\ensuremath{\mathcal{N}}}
\newcommand{\sN}{\ensuremath{\mathbb{N}}}
\newcommand{\cP}{\ensuremath{\mathcal{P}}}
\newcommand{\sP}{\ensuremath{\mathbb{P}}}
\newcommand{\cQ}{\ensuremath{\mathcal{Q}}}
\newcommand{\sQ}{\ensuremath{\mathbb{Q}}}
\newcommand{\cR}{\ensuremath{\mathcal{R}}}
\newcommand{\sR}{\ensuremath{\mathbb{R}}}
\newcommand{\cS}{\ensuremath{\mathcal{S}}}
\newcommand{\sS}{\ensuremath{\mathfrak{S}}}
\newcommand{\cT}{\ensuremath{\mathcal{T}}}
\newcommand{\fT}{\ensuremath{\mathsf{T}}}
\newcommand{\sT}{\ensuremath{\mathbb{T}}}
\newcommand{\cV}{\ensuremath{\mathcal{V}}}
\newcommand{\fV}{\ensuremath{\mathsf{V}}}
\newcommand{\sV}{\ensuremath{\mathbb{V}}}
\newcommand{\sZ}{\ensuremath{\mathbb{Z}}}

@@

\newcommand{\data}{\ensuremath{\sD}}
\newcommand{\guards}[1]{\ensuremath{\sB_{#1}}}
\newcommand{\exprs}[1]{\ensuremath{\sE_{#1}}}
\newcommand{\valuations}[1]{\ensuremath{\sV_{#1}}}
\newcommand{\val}[3][]{%
  \ensuremath{#1{\sigma}^{#2}_{#3}}%
}
\newcommand{\primeit}[1]{#1'}
\newcommand{\doubleprimeit}[1]{#1''}

@@

\noteSBin{%
  Need preliminaries on the universal(?) data domain $\data$
  and valuations.
}

\noteSBin{%
For two valuations $\val{1}{}, \val{2}{} : V \rightarrow \data$,
we denote \[\valdiff{\val{1}{}}{\val{2}{}} \bydef{=} \setdefb{v
  \in V}{\val{1}{}(v) \neq \val{2}{}(v)}\] the set of variables
that are assigned different values by the two valuations.
}

\subsection*{Unifying notations -- remarks}

\todoLHin{proposed notations:}
$g$ are boolean expressions
$e_i$ are updates/assignments

\todoLHin{Remove term algebra or simplify?}
\todoLHin{universal domain $\implies$ remove sort}
\todoLHin{replace assignments by $e_i$}

\noteIn{In Archi we have $a$, whereas in pNets we have $\alpha$, where, approximately $\alpha = 
a(x_1..x_n)$, except that $a$ is port name in archi and action name in pNets.
Is this true? can we unify better? or is it sufficient?
\Ludo

$a$ is a set of ports, but otherwise, yes, this is true.  I think
this is coherent, since, in archi, we have the ``interaction
part'', \ie only ports, plus the data part, \ie variables,
guards, expressions.  In pNets, these are combined.  So, it seems
ok that notations differ\mdash ``uniformisation'' will be
achieved by encoding archi $\rightarrow$ pNets.  \Simon }

@@

@@

\begin{definition}[Components]
  \label{defn:component}
  A \emph{component} is a tuple $(Q, q^0, V, \val{0}{}, P,
  \export{}, \goesto{})$, where
  \begin{itemize}
  \item $Q$ is a set of \emph{states}, with $q^0 \in Q$ the
    \emph{initial state}, 
  \item $V$ is a set of \emph{component variables},
  \item $\val{0}{} : V \rightarrow \data$ is an \emph{initial
    valuation} of the component variables, 
  \item $P$ is a set of \emph{ports}, with
    \noteSB{Added this}\hl{$\export : P \rightarrow 2^V$ 
    the \emph{export function},}
  \item $\goesto{}\, \subseteq
    Q \times 2^P \times \guards{V} \times \exprs{V} \times Q$
%
    is a \emph{transition relation}, with transitions
    labelled by triples consisting of an \emph{interaction}
    $\emptyset \neq a \subseteq P$, a Boolean \emph{guard} $g \in
    \guards{V}$ and an \emph{update expression} $e \in
    \exprs{V}$.
  \end{itemize}
%
  We call \noteSB{Maybe rather $(P, \export)$}\hl{the pair of sets $(V,P)$} the \emph{interface} of the
  component.\,\footnote{%
%
  In practice, component variables are split in two sub-sets:
  local and exported variables.  Only exported variables belong
  to the component interface and can be used for the interactions
  with other components (see \defn{im}).  However, in the context
  of this paper we can simplify by omitting this separation.
%
  }
%
  Notations $q \goesto{a} q'$ and $q \goesto{a}$ are as usual;
  for a component $B$, we denote $Q_B$, $q^0_B$, $P_B$, $V_B$ and
  $\val{0}{B}$ the corresponding constituents of $B$.  We will
  skip the index on the transition relations $\goesto{}$, since it
  is always clear from the context.
\end{definition}

@@

@@

%\smallskip\noindent
\paragraph*{Notations.}
We extensively use indexed structures
over some countable indexed sets, which are equivalent to mappings over
the countable set. % . The indexes will usually be
% integers, bounded or not. Such an indexed family is
%denoted
%follows:
$a_i^{i\in I}$
%, or equivalently  $(i\mapsto a_i)^{i\in I}$
denotes a family of elements $a_i$ indexed over the
set $I$. % Such a family
% is equivalent to the mapping $(i\mapsto a_i)^{i\in I}$.
% To specify the set over which the structure is indexed,
% indexed structures are always denoted with an exponent of the form $i\in I$
% (arithmetic only appears in the indexes if necessary).
$a_i^{i\in I}$ defines both $I$ the set over which the family is
indexed (called \emph{range}), and $a_i$ the elements of the family.
E.g., $a^{i\in\{3\}}$ is the mapping with a single entry $a$ at index
$3$ ; abbreviated $(3\mapsto a)$ in the following.
When this is not
ambiguous, we shall use notations for sets, and typically write
``indexed set over I'' when formally we should speak of multisets, and
write $x\in a_i^{i\in I}$ to mean $\exists i\in I.\, x=a_i$.  An empty
family is denoted $\emptyset$. We
denote classically with an overline -- $\overline{a}$  -- a family when the indexing set 
is
not meaningful.  $\uplus$ is the disjoint union on
indexed sets.

\paragraph*{Term algebra.}
Our models rely on a notion of parameterised actions, that are
symbolic expressions using data types and variables. As our model aims
at encoding the low-level behaviour of possibly very different
programming languages, we do not want to impose one specific algebra
for denoting actions, nor any specific communication mechanism. So we
leave unspecified the constructors of the algebra that will allow building
expressions and actions. Moreover, we use a generic {\em action interaction}
mechanism, based on (some sort of) unification between two or more action
expressions, to express various kinds of communication or
synchronisation mechanisms.

\def\Talg{\mathcal{T}_{\Sigma,\P}}
Formally, we assume the existence of a term algebra $\Talg$,
where $\Sigma$ is the signature of the data and action constructors,
and $\P$ a set of variables. Within $\Talg$, we distinguish a set of
data expressions $\mathcal{E}_\P$, including a set of boolean
expressions \todoLHin{$\mathcal{B} \to g$}
$\mathcal{B}_{\P}$ ($\mathcal{B}_{\P}\subseteq\mathcal{E}_\P$).
On top of $\mathcal{E}_\P$ we build the action algebra
$\mathcal{A}_\P$, with $\mathcal{A}_P\subseteq\mathcal{T}_\P,
\mathcal{E}_P\cap\mathcal{A}_P=\emptyset$;
naturally action terms will use data expressions as subterms.
To be able to reason about the data flow between pLTSs, we
distinguish \emph{input variables} of the form $?x$ within terms; the function
$\vars(t)$ identifies the set of variables in a term
$t\in\AlgT$, and $iv(t)$ returns its input variables.
Action algebras can encode naturally usual point-to-point message passing calculi (using 
$a(?x_1,...,?x_n)$ for inputs, $a(v_1,..,v_n)$ for outputs), but it also allows
for more general mechanisms, like gate negociation in Lotos, or broadcast
communications.

>>>>

\newcommand{\noteSB}[2][color=green!40, size=\tiny]{\todo[#1]{{#2}\Simon}}
\newcommand{\noteSBin}[2][inline,color=green!40]{\todo[#1]{{#2}\Simon}}
\newcommand{\todoSB}[2][color=green!40, size=\tiny]{\todo[#1]{\textbf{To-do Simon:} {#2}}}
\newcommand{\todoSBin}[2][inline,color=green!40]{\todo[#1]{\textbf{To-do Simon: } {#2}}}
\newcommand{\Ludo}{\\\hfill\mdash Ludo}
\newcommand{\noteLH}[2][color=orange!40, size=\tiny]{\todo[#1]{{#2}\Ludo}}
\newcommand{\noteLHin}[2][inline,color=orange!40]{\todo[#1]{{#2}\Ludo}}
\newcommand{\todoLH}[2][color=orange!40, size=\tiny]{\todo[#1]{\textbf{To-do Ludo:} {#2}}}
\newcommand{\todoLHin}[2][inline,color=orange!40]{\todo[#1]{\textbf{To-do Ludo: } {#2}}}

@@

\newcommand{\add}[2][Added]{\todo[color=blue!20, size=\tiny]{#1}{\color{blue}#2}}
%% \newcommand{\remove}[2][Removed]{\todo[color=red!20, size=\tiny]{#1}}
\newcommand{\remove}[2][Removed]{\todo[color=red!20, size=\tiny]{#1}{\color{red}\st{#2}}}
%% \newcommand{\replace}[3][Replaced]{\todo[color=blue!20, size=\tiny]{#1}{\color{blue}#3}}
\newcommand{\replace}[3][Replaced]{\todo[color=blue!20, size=\tiny]{#1}{\color{blue}#3}{\color{red}\st{#2}}}

\newcommand{\addSB}[1]{\add[Added by Simon]{#1}}
\newcommand{\removeSB}[1]{\remove[Removed by Simon]{#1}}
\newcommand{\replaceSB}[2]{\replace[Replaced by Simon]{#1}{#2}}

@@

\newcommand{\cA}{\ensuremath{\mathcal{A}}}
\newcommand{\fA}{\ensuremath{\mathsf{A}}}
\newcommand{\sA}{\ensuremath{\mathbb{A}}}
\newcommand{\cB}{\ensuremath{\mathcal{B}}}
\newcommand{\fB}{\ensuremath{\mathsf{B}}}
\newcommand{\sB}{\ensuremath{\mathbb{B}}}
\newcommand{\cC}{\ensuremath{\mathcal{C}}}
\newcommand{\sC}{\ensuremath{\mathbb{C}}}
\newcommand{\cD}{\ensuremath{\mathcal{D}}}
\newcommand{\fD}{\ensuremath{\mathsf{D}}}
\newcommand{\sD}{\ensuremath{\mathbb{D}}}
\newcommand{\cE}{\ensuremath{\mathcal{E}}}
\newcommand{\sE}{\ensuremath{\mathbb{E}}}
\newcommand{\cF}{\ensuremath{\mathcal{F}}}
\newcommand{\cG}{\ensuremath{\mathcal{G}}}
\newcommand{\cH}{\ensuremath{\mathcal{H}}}
\newcommand{\cI}{\ensuremath{\mathcal{I}}}
\newcommand{\sI}{\ensuremath{\mathbb{I}}}
\newcommand{\cM}{\ensuremath{\mathcal{M}}}
\newcommand{\cN}{\ensuremath{\mathcal{N}}}
\newcommand{\sN}{\ensuremath{\mathbb{N}}}
\newcommand{\cP}{\ensuremath{\mathcal{P}}}
\newcommand{\sP}{\ensuremath{\mathbb{P}}}
\newcommand{\cQ}{\ensuremath{\mathcal{Q}}}
\newcommand{\sQ}{\ensuremath{\mathbb{Q}}}
\newcommand{\cR}{\ensuremath{\mathcal{R}}}
\newcommand{\sR}{\ensuremath{\mathbb{R}}}
\newcommand{\cS}{\ensuremath{\mathcal{S}}}
\newcommand{\sS}{\ensuremath{\mathfrak{S}}}
\newcommand{\cT}{\ensuremath{\mathcal{T}}}
\newcommand{\fT}{\ensuremath{\mathsf{T}}}
\newcommand{\sT}{\ensuremath{\mathbb{T}}}
\newcommand{\cV}{\ensuremath{\mathcal{V}}}
\newcommand{\fV}{\ensuremath{\mathsf{V}}}
\newcommand{\sV}{\ensuremath{\mathbb{V}}}
\newcommand{\sZ}{\ensuremath{\mathbb{Z}}}

@@

\newcommand{\ordbool}{\ensuremath{\sB^{<}}}
\newcommand{\data}{\ensuremath{\sD}}
\newcommand{\signature}{\ensuremath{\Sigma}}
\newcommand{\variables}{\ensuremath{\cV}}
\newcommand{\Talg}{\ensuremath{\cT_{\signature,\variables}}}
\newcommand{\actions}[1]{\ensuremath{\cA_{#1}}}
\newcommand{\guards}[1]{\ensuremath{\ordbool_{#1}}}
\newcommand{\exprs}[1]{\ensuremath{\sE_{#1}}}
\newcommand{\assigns}[1]{\ensuremath{\sA_{#1}}}
\newcommand{\valuations}[1]{\ensuremath{\data^{#1}}}
\newcommand{\val}[3][]{\ensuremath{#1{\sigma}^{#2}_{#3}}}

\newcommand{\primeit}[1]{#1'}
\newcommand{\doubleprimeit}[1]{#1''}

@@

\section{Preliminaries}
\label{secn:preliminaries}

\subsection{Notations}
\label{secn:notations}

We extensively use \replaceSB{indexed structures
over some countable indexed sets, which are equivalent to mappings over
the countable set. }{structures indexed over some finite or countable sets of indices, which can be equivalently defined as mappings over such index sets. }% . The indexes will usually be
% integers, bounded or not. Such an indexed family is
%denoted
%follows:
\addSB{Thus, }
$a_i^{i\in I}$
%, or equivalently  $(i\mapsto a_i)^{i\in I}$
denotes a family of elements $a_i$ indexed over the
set $I$. % Such a family
% is equivalent to the mapping $(i\mapsto a_i)^{i\in I}$.
% To specify the set over which the structure is indexed,
% indexed structures are always denoted with an exponent of the form $i\in I$
% (arithmetic only appears in the indexes if necessary).
\replaceSB{$a_i^{i\in I}$ }{This notation }defines both $I$ the set over which the family is
indexed (called \emph{range}), and $a_i$ the elements of the family.
E.g., $a^{i\in\{3\}}$ is the mapping with a single entry $a$ at index
$3$ ; abbreviated $(3\mapsto a)$ in the following.
When this is not
ambiguous, we shall use notations for sets, and typically write
``indexed set over I''\replaceSB{ when}{, even though} formally we should speak of multisets; and
write $x\in a_i^{i\in I}$ to mean $\exists i\in I.\, x=a_i$.  An empty
family is denoted $\emptyset$. \replaceSB{
We denote classically with an overline -- $\overline{a}$ -- a family when the indexing set 
is
not meaningful.}{We use a classical overline notation, \eg $\overline{a}$, for a family, where the indexing set is irrelevant} \replaceSB{$\uplus$ is }{Finally, we use $\uplus$ to denote }the disjoint union on
indexed sets.

@@

\subsection{Term algebra}
\label{secn:terms}

\todoLHin{Remove term algebra or simplify?}
\todoLHin{universal domain $\implies$ remove sort}
\todoLHin{replace assignments by $e_i$

  In fact, in the pNet action terms, you need expressions, not
  assignements.  Therefore assuming that all expressions are
  assignments will not work, so the right approach is to have
  expressions (for pNets) and assignments (for BIP) separately.
  \Simon
%
}

\noteSBin{In Archi we have $a$, whereas in pNets we have $\alpha$, where, approximately $\alpha = 
a(x_1..x_n)$, except that $a$ is port name in archi and action name in pNets.
Is this true? can we unify better? or is it sufficient?
\Ludo

$a$ is a set of ports, but otherwise, yes, this is true.  I think
this is coherent, since, in archi, we have the ``interaction
part'', \ie only ports, plus the data part, \ie variables,
guards, expressions.  In pNets, these are combined.  So, it seems
ok that notations differ\mdash ``uniformisation'' will be
achieved by encoding archi $\rightarrow$ pNets.
}

\replaceSB{Our models rely }{The pNet model in \secn{pNets} relies }on a notion of parameterised actions, that are
symbolic expressions using data types and variables. As \replaceSB{our model aims }{we aim }at encoding the low-level behaviour of possibly very different
programming languages, we do not want to impose one specific algebra
for denoting actions, nor any specific communication mechanism. So we
leave unspecified the constructors of the algebra that will allow building
expressions and actions. Moreover, we use a generic {\em action interaction}
mechanism, based on (some sort of) unification between two or more action
expressions, to express various kinds of communication or
synchronisation mechanisms.

%\def\Talg{\mathcal{T}_{\Sigma,\P}}  % Moved to the header

\noteSBin{proposed notations:

  $g$ are boolean expressions;
  $e_i$ are updates/assignments
  \Ludo

  In fact, I am not sure whether I have correctly understood this
  suggestion: do you mean, \eg using $g$ to denote the set of
  Boolean expressions or individual Boolean expressions? Using
  lower case letters to denote sets might be problematic, since
  we (at least I) use them a lot to denote instances, as in
  \[q \goesto{a, g, e} q'\,.\]
  In fact, I even use capital letters: in the definition of
  interaction model composition, I use
  \[\cdot \goesto{a, G, E} \cdot\]
  in the conclusion of the rule.  However, for this latter I can
  easily change the notation.

  I have set up macros for, general and Boolean expressions,
  actions, variables \etc[,] which I have substituted everywhere,
  where appropriate in the following paragraph.
}

  \begin{tabular}{
      @{}p{0.15\columnwidth}@{}
      p{0.35\columnwidth}@{}
      p{0.5\columnwidth}@{}
    }
    {\bf Output} & {\bf Command} & {\bf Definition}
    \\
    $\data$ &
    {\ttfamily\textbackslash data} &
    {\ttfamily\textbackslash sD}
    \\
    \signature &
    {\ttfamily\textbackslash signature} &
    {\ttfamily\textbackslash Sigma}
    \\
    \variables &
    {\ttfamily\textbackslash variables} &
    {\ttfamily\textbackslash cV}
    \\
    \Talg &
    {\ttfamily\textbackslash Talg} &
    {\ttfamily\textbackslash cT\_\{\textbackslash signature,\textbackslash variables\}}
    \\
    \actions{\#1} &
    {\ttfamily\textbackslash actions\{\#1\}} &
    {\ttfamily\textbackslash cA\_\{\#1\}}
    \\
    \guards{\#1} &
    {\ttfamily\textbackslash guards\{\#1\}} &
    {\ttfamily\textbackslash sB\^{}\{<\}\{\#1\}}
    \\
    \exprs{\#1} &       
    {\ttfamily\textbackslash exprs\{\#1\}} &
    {\ttfamily\textbackslash sE\_\{\#1\}}
    \\
    \assigns{\#1} &
    {\ttfamily\textbackslash assigns\{\#1\}} &
    {\ttfamily\textbackslash sA\_\{\#1\}}
    \\
    \valuations{\#1} &
    {\ttfamily\textbackslash valuations\{\#1\}} &
    {\ttfamily\textbackslash data\^{}\{\#1\}}
    \\[6pt]
    \val{\#2}{\#3} &
    {\ttfamily\textbackslash val\{\#2\}\{\#3\}} &
    {\ttfamily\#1\{\textbackslash sigma\}\^{}\{\#2\}\_\{\#3\}}
    \\[6pt]
    \val[\tilde]{\#2}{\#3} &
    {\ttfamily\textbackslash val[\textbackslash tilde]\{\#2\}\{\#3\}} &
    {\ttfamily\#1\{\textbackslash sigma\}\^{}\{\#2\}\_\{\#3\}}
  \end{tabular}
\noteSBin{commands above (tabular generates an error inside a
  note)}

Formally, we assume the existence of a term algebra $\Talg$,
where $\signature$ is the signature of the data and action constructors,
and $\variables$ a set of \emph{variables}. Within $\Talg$, we distinguish a set of
\emph{data expressions} $\exprs{\variables}$ \replaceSB{, including}{and} a set of \emph{Boolean
expressions} \todoLH{$\mathcal{B} \to g$}
$\guards{\variables}$ ($\guards{\variables}\subseteq\exprs{\variables}$).
On top of $\exprs{\variables}$ we build the \emph{action algebra}
$\actions{\variables}$, with $\actions{\variables}\subseteq\Talg$ \replaceSB{,}{and}
$\exprs{\variables}\cap\actions{\variables} = \emptyset$;
naturally action terms will use data expressions as subterms.
\addSB{In addition to the above notations, we will use
\[
\assigns{\variables} \bydef{=}
\Setdef{(x_i := e_i)^{i \in I}}{
  x_i^{i \in I}\subseteq \variables,
  e_i^{i \in I}\subseteq \exprs{\variables},
  I \text{ is a finite index set}
}
\]
to denote the set of \emph{variable assignments}.}

To be able to reason about the data flow between pLTSs, we
distinguish \emph{input variables} of the form $?x$ within terms; the function
$\vars(t)$ identifies the set of variables in a term
$t\in\Talg$, and $iv(t)$ returns its input variables.
Action algebras can \replaceSB{encode naturally }{naturally encode }usual point-to-point message passing calculi (using 
$a(?x_1,\dots,?x_n)$ for inputs, $a(v_1,\dots,v_n)$ for outputs), but it also allows
for more general mechanisms, like gate negotiation in Lotos, or broadcast
communications.

\addSB{%
%
  Finally, we assume the existence of a universal data domain
  given as a partially-ordered set $(\data, \order)$, potentially
  encompassing several copies of any given data type with
  different orders.  For example, we assume that $(\data,
  \order)$ comprises both the unordered set of Booleans $\sB =
  (\{\true, \false\}, \emptyset)$ and the naturally ordered one
  $\sB^< = (\{\true, \false\}, \{\false < \true\})$, and
  similarly for integer and real numbers; as well as the set of
  intervals ordered by inclusion.

  For a set of variables $V \subseteq \variables$, we denote
  $\valuations{V} \bydef{=} \{\val{}{}: V \rightarrow \data\}$
  the set of \emph{valuations} of the variables in $V$.
  Valuations extend canonically to expressions.  Thus
  $\val{}{}(x)$ and $\val{}{}(e)$ denote, respectively, the
  values of the variable $x$ and the expression $e$ under the
  valuation $\val{}{}$.  For a valuation $\val{}{} \in
  \valuations{V}$ and an assignment $(x_i := e_i)^{i \in I} \in
  \assigns{V}$, we denote $\val{}{}[(x_i := e_i)^{i \in I}]$ the
  valuation defined by
  \[
  \val{}{}\bigl[(x_i := e_i)^{i \in I}\bigr](x) \bydef{=}
  \begin{cases}
    \val{}{}(x), & \text{if } x \not\in x_i^{i \in I}\,,\\
    \val{}{}(e_i), & \text{if } x = x_i, \text{for some } i \in I\,.
  \end{cases}
  \]
  For two valuations $\val{1}{}, \val{2}{} : V \rightarrow
  \data$, we denote \[\valdiff{\val{1}{}}{\val{2}{}} \bydef{=}
  \setdefb{v \in V}{\val{1}{}(v) \neq \val{2}{}(v)}\] the set of
  variables that are assigned different values by the two
  valuations.  As usual, we write $\val{1}{} \order \val{2}{}$
  iff $\val{1}{}(x) \order \val{2}{}(x)$, for all $x \in V$.  An
  expression $e \in \exprs{V}$ is called \emph{monotonic} if, for
  any two valuations $\val{1}{}, \val{2}{} \in \valuations{V}$,
  $\val{1}{} \order \val{2}{}$ implies $\val{1}{}(e) \order
  \val{2}{}(e)$.
%
}

@@

\begin{definition}[Components]
  \label{defn:component}
  A \emph{component} is a tuple $(Q, q^0, V, \val{0}{}, P,
  \export{}, \goesto{})$, where
  \begin{itemize}
  \item $Q$ is a set of \emph{states}, with $q^0 \in Q$ the
    \emph{initial state}, 
  \item $V$ is a set of \emph{component variables},
  \item $\val{0}{} : V \rightarrow \data$ is an \emph{initial
    valuation} of the component variables, 
  \item $P$ is a set of \emph{ports}, with
    \noteSB{Added this}\hl{$\export : P \rightarrow 2^V$ 
    the \emph{export function},}
  \item $\goesto{}\, \subseteq
    Q \times 2^P \times \guards{V} \times \exprs{V} \times Q$
%
    is a \emph{transition relation}, with transitions
    labelled by triples consisting of an \emph{interaction}
    $\emptyset \neq a \subseteq P$, a monotonic Boolean \emph{guard} $g \in
    \guards{V}$ and a monotonic \emph{update expression} \todoSB{$e \rightarrow u$}\hl{$e \in
    \exprs{V}$}.
  \end{itemize}
%
  We call \noteSB{Maybe rather $(P, \export)$}\hl{the pair of sets $(V,P)$} the \emph{interface} of the
  component.\,\footnote{%
%
  In practice, component variables are split in two sub-sets:
  local and exported variables.  Only exported variables belong
  to the component interface and can be used for the interactions
  with other components (see \defn{im}).  However, in the context
  of this paper we can simplify by omitting this separation.
%
  }
%
  Notations $q \goesto{a} q'$ and $q \goesto{a}$ are as usual;
  for a component $B$, we denote $Q_B$, $q^0_B$, $P_B$, $V_B$ and
  $\val{0}{B}$ the corresponding constituents of $B$.  We will
  skip the index on the transition relations $\goesto{}$, since it
  is always clear from the context.
\end{definition}

@@

%****************************************************************
\subsection{Graphical representation of composed systems}
\label{secn:connectors}

@@
